\chapter{Theoretical Foundations}

This chapter develops the theoretical machinery the system in Chapter~3 is
built on: the LLM architectures used, the three families of evaluation
metrics, the factuality methodology, the content-extraction stack, and the
browser-extension framework. Each section ties the theory to a concrete
choice in the system.

\section{Large Language Models}

\subsection{The Transformer Architecture}

The Transformer \cite{Vaswani2017} replaced recurrent and convolutional
sequence models for natural-language tasks by relying entirely on attention.
For a query $Q$, key $K$, and value $V$, the scaled dot-product attention is

\[
\text{Attention}(Q,K,V) \;=\; \text{softmax}\!\left(\frac{QK^{\top}}{\sqrt{d_k}}\right) V,
\]

\noindent where $d_k$ is the key dimension. \textit{Multi-head} attention
projects $Q, K, V$ into $h$ subspaces, runs attention in each, and
concatenates the results, allowing the model to attend to different
positional relationships simultaneously.

\subsection{Architectural Families}

Modern LLMs fall into three architectural families.

\begin{itemize}
  \item \textit{Encoder-only} models such as BERT and PhoBERT \cite{Nguyen2020}
        produce contextual embeddings for input tokens; they are used in this
        thesis as the embedding backbone for BERTScore (Section~2.2.3).
  \item \textit{Decoder-only} models such as GPT, Claude, Gemini, and
        PhoGPT \cite{Phan2023} generate text autoregressively and are used
        as the proposers and aggregator of the MoA pipeline (Chapter~3).
  \item \textit{Encoder--decoder} models such as T5 and
        ViT5 \cite{Nguyen2022} encode the input and condition decoder
        generation on it; ViT5 is used as the routing fallback for short
        Vietnamese inputs.
\end{itemize}

\subsection{Models Used in the System}

\begin{itemize}
  \item \textbf{GPT-4o} (OpenAI, 2024-08-06 snapshot) --- aggregator in the
        MoA pipeline and the strongest-single-model baseline.
  \item \textbf{GPT-4o-mini} (OpenAI, 2024-07-18 snapshot) --- proposer
        and judge model; default LLM-judge in Axis~B.
  \item \textbf{Claude Haiku 4.5} (Anthropic) --- proposer; included to
        widen the proposer family beyond OpenAI.
  \item \textbf{Gemini 2.5 Flash} (Google) --- proposer; the third
        proposer family used in MoA.
  \item \textbf{ViT5-large} (VinAI) --- routing fallback for short
        Vietnamese inputs in the complexity-based routing layer.
  \item \textbf{PhoBERT-base} (\texttt{vinai/phobert-base}) --- contextual
        encoder for BERTScore.
  \item \textbf{PhoGPT-4B-Chat} (VinAI) --- evaluated for on-device
        deployment; not used in the headline experiments due to GPU
        constraints discussed in Chapter~4.
\end{itemize}

\subsection{Reasoning-Model Considerations}

Reasoning-tuned models such as OpenAI \texttt{o4-mini} and
\texttt{o3-mini} require special handling: temperature, top-$p$, and
penalty parameters are unsupported, and the streaming protocol differs from
chat-completion models. The service layer in Chapter~3 abstracts these
differences so that all 14 supported models can be selected interchangeably
at runtime.

\section{Summarization Evaluation Metrics (Axis A)}

\subsection{ROUGE}

ROUGE \cite{Lin2004} (Recall-Oriented Understudy for Gisting Evaluation)
measures $n$-gram and longest-common-subsequence overlap between a candidate
summary $C$ and a reference $R$. Let $g_n$ denote the set of $n$-grams.

\[
\text{ROUGE-}n_{\text{recall}} \;=\; \frac{|g_n(C) \cap g_n(R)|}{|g_n(R)|},
\qquad
\text{ROUGE-}n_{\text{prec}} \;=\; \frac{|g_n(C) \cap g_n(R)|}{|g_n(C)|},
\]

\noindent and ROUGE-$n_{F1}$ is the harmonic mean of the two.
ROUGE-1 and ROUGE-2 use unigrams and bigrams respectively; ROUGE-L is
based on the longest common subsequence and rewards in-order overlap.

\textbf{Caveat for Vietnamese.} ROUGE is computed on whitespace tokens.
A Vietnamese word such as \textit{kiểm chứng} is two tokens, and a paraphrase
that uses \textit{thẩm định} instead receives zero credit even though the
meaning is identical.

\subsection{BLEU}

BLEU is a precision-oriented metric originally designed for machine
translation. For an $n$-gram set, modified precision $p_n$ clips counts at
the maximum reference count; BLEU then combines $p_1, \ldots, p_4$ as a
weighted geometric mean and multiplies by a brevity penalty $BP$:

\[
\text{BLEU} \;=\; BP \cdot \exp\!\left(\sum_{n=1}^{4} w_n \log p_n \right),
\qquad
BP \;=\; \min\!\left(1, \exp\!\left(1 - \tfrac{r}{c}\right)\right),
\]

\noindent where $r$ is the reference length, $c$ is the candidate length,
and $w_n = 1/4$. BLEU shares ROUGE's tokenization fragility for Vietnamese.

\subsection{BERTScore (with PhoBERT)}

BERTScore \cite{Zhang2020} replaces $n$-gram matching with cosine similarity
between contextual embeddings. For each token $c_i$ in the candidate, the
best-matching token $r_j$ in the reference contributes
$\cos(\mathbf{e}(c_i), \mathbf{e}(r_j))$ to BERTScore-Precision; recall is
defined symmetrically. The final score $F_{\text{BERT}}$ is the harmonic
mean.

In this thesis, the embedding model is \texttt{vinai/phobert-base}; inputs
are truncated to $256$ tokens before scoring, which both fits the model's
context window and standardizes the cost of an evaluation call.
BERTScore captures semantic similarity that overlap metrics miss, but it
still scores against the source article, so it inherits the same
\textit{content-retention} interpretation as ROUGE and BLEU.

\subsection{Compression Rate}

Compression rate is $|S| / |D|$ expressed as a percentage, where $|S|$ and
$|D|$ are the token counts of the summary and the source. It is not a
quality metric, but it explains \textit{why} overlap metrics behave as they
do: an editorial summary at $30\%$ compression has fewer surface forms to
match than a draft-stitching summary at $50\%$, and overlap will fall
even when quality rises.

\subsection{Methodological Limit of Axis A}

The most important caveat of this thesis, restated formally:
\textit{ROUGE, BLEU, and BERTScore computed against the source article
measure content retention, not summary quality}. When the system is
allowed (or encouraged) to paraphrase, an editorial rewrite that improves
clarity will reduce token overlap. Wang et al.\ \cite{Wang2024} bypass
this issue by using an LLM judge as ground truth; the same choice is made
in Axis~B of this thesis.

\section{LLM-as-Judge Methodology (Axis B)}

\subsection{Rubric Evaluation (FLASK-Derived)}

A rubric judge scores a summary along several named dimensions
on an integer scale. Following FLASK \cite{Ye2023}, this thesis uses five
dimensions on a 1--5 scale: \textit{faithfulness}, \textit{coverage},
\textit{fluency}, \textit{conciseness}, and a holistic \textit{overall}.
The judge model receives the source article, the candidate summary, and a
Vietnamese-journalism rubric specification, and is instructed to return a
JSON object with the five scores and a one-sentence justification per
dimension.

\subsection{Absolute Scoring (MT-Bench Style)}

The MT-Bench protocol \cite{Zheng2023} asks the judge for a single
holistic integer score in $\{1,\ldots,10\}$. This is faster than the
rubric and is included as a secondary signal; it does not replace the
rubric.

\subsection{Pairwise Preference (AlpacaEval Style)}

A pairwise judge is shown two candidate summaries $A$ and $B$ for the same
source and is asked to pick a winner (or a tie). The empirical headline of
this thesis is reported in pairwise form because pairwise judges show
higher inter-judge agreement than absolute judges. Two protocol details
matter:

\begin{itemize}
  \item \textit{Position randomization.} LLM judges have a documented bias
        toward the first-presented candidate. Every pairwise call randomly
        assigns the two summaries to positions $A$ and $B$, and the
        position mapping is recorded so the winning candidate is correctly
        attributed.
  \item \textit{Sign-test significance.} For $w$ wins out of $n$ decisive
        verdicts (ties excluded), the two-sided sign-test $p$-value is

        \[
        p \;=\; 2 \cdot \Pr\!\left[\,X \geq \max(w, n - w) \;\mid\; X \sim \text{Bin}(n, 0.5)\right],
        \]

        \noindent which is the probability of an outcome at least as
        extreme as the observed one under $H_0: P(\text{A wins}) = 0.5$.
        Throughout this thesis, $p < 0.05$ is treated as the threshold
        for ``unlikely to be chance.''
\end{itemize}

\subsection{Length-Controlled Win Rate (Dubois 2024)}

LLM judges prefer longer responses \cite{Dubois2024}. To control for this,
pairwise verdicts are bucketed by the length ratio $|A|/|B|$:

\begin{itemize}
  \item $|A|/|B| < 0.85$ --- $A$ is much shorter;
  \item $0.85 \leq |A|/|B| \leq 1.15$ --- comparable length;
  \item $|A|/|B| > 1.15$ --- $A$ is much longer.
\end{itemize}

\noindent The length-bucketed win rate averages the within-bucket win rates
of buckets that meet a minimum count ($\text{MIN\_BUCKET\_N} = 5$). If
raw and bucketed rates agree, length bias is unlikely to explain the
result.

\section{Factuality Evaluation}

Factuality is treated as a per-claim binary problem. Given a summary $S$
and source $D$:

\begin{enumerate}
  \item \textit{Claim splitting.} Decompose $S$ into atomic claims
        $\{c_1, \ldots, c_k\}$ via a structured-output LLM call.
  \item \textit{Entailment classification.} For each $c_j$, classify into
        $\{\,\text{entailed},\, \text{contradicted},\, \text{not mentioned}\,\}$
        relative to $D$, using \texttt{gpt-4o-mini} with a constrained JSON
        schema.
  \item \textit{Aggregation.} Report the entailment ratio
        $|\{c_j : \text{entailed}\}| / k$, the count of contradicted claims
        (hallucinations), and store contradicting claims verbatim for
        manual inspection.
\end{enumerate}

A summary with entailment ratio $> 95\%$ and zero hallucinations is
considered factually faithful at the per-claim level; the experiment in
Chapter~4 reports both the average and worst-case values per approach.

\section{Human Evaluation \& Inter-rater Agreement (Axis C)}

\subsection{Blind K-Way Ranking}

Axis~C uses blind $K$-way ranking. For each article, $K = 3$ candidate
summaries (fused, GPT-4o-alone, and a single proposer draft) are presented
to the rater as \textit{Bản A, Bản B, Bản C}, with the model identities
hidden. The rater drags the candidates into a total order from best to
worst and writes a one-sentence Vietnamese justification per candidate.
The interface deliberately mirrors Chatbot Arena
\cite{Zheng2023} but for $K = 3$.

\subsection{Fleiss' Kappa}

Fleiss' $\kappa$ \cite{Landis1977} extends Cohen's $\kappa$ to
$\geq 3$ raters. For $N$ items, $n$ raters per item, and $k$ categories,
$\kappa = (P - P_e) / (1 - P_e)$, where $P$ is the average per-item
agreement (over rater pairs) and $P_e$ is the chance-expected agreement
under marginal category frequencies.

The Landis--Koch interpretation bands used in this thesis are:

\begin{tabular}{ll}
$\kappa < 0$       & worse than chance \\
$0.00$--$0.20$     & slight \\
$0.21$--$0.40$     & fair \\
$0.41$--$0.60$     & moderate \\
$0.61$--$0.80$     & substantial \\
$0.81$--$1.00$     & almost perfect \\
\end{tabular}

For ranking data, $\kappa$ is computed on per-position labels: for each
$(item, rank)$ pair, the ``category'' assigned by a rater is the candidate
that received that rank. The implementation is described in
Section~3.6.4.

\section{Content Extraction}

The browser extension is invoked on arbitrary news pages; before the article
can be summarized, the system must isolate the article body from
navigation, advertisements, comments, and related-article rails.

\textbf{Mozilla Readability + JSDOM.} The backend reuses
\textit{Mozilla Readability} --- the algorithm behind Firefox's Reader View
--- run inside a JSDOM environment. Readability scores DOM nodes by
content-density heuristics and returns the highest-scoring subtree as the
main article.

\textbf{Vietnamese-news edge cases.} Vietnamese outlets use several
DOM patterns that defeat the default heuristics:

\begin{itemize}
  \item \textit{VnExpress} embeds article paragraphs inside a custom
        web component with a Shadow DOM root; the extension must pierce
        the shadow boundary on the client before sending text.
  \item \textit{Cloudflare-protected sites} (\textit{laodong.vn},
        \textit{nld.com.vn}) require the extraction to happen in the
        content script, where the browser session already holds the
        Cloudflare cookie.
  \item \textit{VietnamNet} renders the article through a server-side
        template that injects ad slots inline with paragraphs; the
        Readability score must ignore script-tagged ad blocks.
\end{itemize}

\section{Search-Augmented Generation for Fact-Checking}

\textbf{Tavily API.} Tavily is a search API tuned for LLM consumption:
each result is a tuple \texttt{(title, url, snippet)} with a relevance
score. The fact-check service issues a single Tavily query per candidate
claim, requests $6$ results, and feeds the snippets to GPT-4o-mini for
verdict prediction.

\textbf{RAG framing.} The pipeline is a one-shot retrieval-augmented
generation (RAG) instance: the source article and Tavily snippets are the
``retrieved context'' for a verdict LLM. The thesis does not introduce a
vector database; live web evidence is fresher than any cached corpus and is
sufficient for breaking-news verification.

\textbf{Trust score.} The verdict LLM returns a trust score in $[0, 100]$,
a boolean \texttt{verified} flag (true iff trust $\geq 75$), and a list of
source URLs. The extension renders these in a fact-check modal next to the
summary.

\section{Browser Extension Development}

\subsection{Chrome Extension Manifest V3}

Manifest V3 (MV3) is the current Chromium extension specification. The
relevant constraints are:

\begin{itemize}
  \item \textit{Service worker} replaces the persistent background page;
        any cross-tab state must be persisted to \texttt{chrome.storage}
        and re-hydrated on each event.
  \item \textit{Permissions model} requires explicit host permissions for
        every site the extension can read; this is the reason the
        supported-domain list is enumerated rather than wildcarded.
  \item \textit{Content scripts} run in the page world; CSS is namespaced
        via Shadow DOM to avoid clashing with the host page's styles.
\end{itemize}

\subsection{Plasmo Framework}

\textit{Plasmo} is a React + TypeScript framework for browser extensions
that provides hot-reload during development, a unified configuration
file, and automatic build outputs for Chromium and Firefox. The Fiber
extension uses Plasmo to compose its content script (sidebar),
popup, and background service worker from a single TypeScript codebase.

\section{Summary}

This chapter laid out the LLM architectures, the three evaluation families
(overlap, LLM-judge, human), the factuality methodology, the
content-extraction stack, and the browser-extension framework that the
system in Chapter~3 builds on. The crucial theoretical claim of the
chapter is that overlap metrics computed against a source article measure
content retention, not summary quality; this claim, made formal in
Section~2.2.5, motivates the addition of Axis~B and Axis~C in the
evaluation framework.
