\section{Summary}
\hspace{1cm} This chapter presented TOCL+, our extension to OCL for specifying and 
verifying temporal properties in UML models. We began by examining the limitations 
of standard OCL and existing temporal extensions, particularly the inability to 
express event-based and bounded existence properties. To address these limitations, 
we introduced TOCL+, which extends TOCL with two key event constructs: 
\texttt{isCalled} for detecting operation invocations and \texttt{becomesTrue} for 
capturing state changes. We complemented these with bounded existence operators that 
enable specifying constraints on event frequencies. We provided formal semantics for 
these constructs, defining them precisely in terms of state transitions, and developed 
a complete EBNF grammar to support automated processing. For verification, we presented 
a three-step approach: first transforming the application model into a filmstrip model, 
then translating TOCL+ specifications into standard OCL constraints over this model, 
and finally using the USE Model Validator to check these constraints. We demonstrated 
the effectiveness of our approach on a Software System example, showing how TOCL+ 
can express and verify complex temporal properties including safety, liveness, and 
fairness constraints that were previously inexpressible in OCL. This combination of 
enhanced expressiveness with automated verification provides modelers with powerful 
new capabilities for ensuring the correctness of dynamic behavior in their UML models.