%%% Version 1
As established in Chapter 1, OCL provides robust support for specifying structural properties 
and simple behavioral constraints in UML models. However, OCL has significant limitations 
when applied to dynamic system behavior. It operates on single system states or individual 
transitions, making it unable to express properties that span across multiple states or 
respond to events within the system's execution. These limitations become particularly 
problematic for modern software systems, which frequently require complex temporal and 
reactive behaviors.

This chapter presents two main contributions to address these limitations:

First, we present TOCL+, a comprehensive extension of OCL that enhances its expressiveness 
for dynamic system aspects. TOCL+ combines temporal operators adapted from Temporal OCL 
research with novel event-based constructs. The temporal operators enable reasoning about 
system evolution over time with constructs like \textit{always}, \textit{sometime}, and 
\textit{until}. Our event constructs address a critical gap by enabling the detection of 
specific occurrences during system execution, such as operation calls and state changes. 
Together, these extensions create a more powerful specification language capable of expressing 
complex dynamic requirements such as "when a login attempt fails three consecutive times, 
the account must be locked."

Second, we introduce a transformation approach that enables the verification of TOCL+ 
specifications using existing model checking tools. This approach transforms UML/OCL models 
into filmstrip models that expose state sequences, and translates TOCL+ specifications into 
standard OCL constraints that can be verified within the filmstrip context. This transformation 
bridges the gap between expressing temporal requirements and verifying them, providing a 
complete solution that integrates with established verification technologies.

The chapter is organized as follows:
\begin{itemize}
    \item Section 2.2 presents the TOCL+ language extension, covering both temporal 
    specification capabilities and our novel event-based constructs, as well as their 
    integration.
    
    \item Section 2.3 details the transformation approach for verification, explaining 
    how UML/OCL models are transformed to filmstrip models and how TOCL+ specifications 
    are translated to standard OCL for verification.
\end{itemize}

By addressing both specification and verification aspects, this chapter provides a 
comprehensive solution to the challenge of expressing and verifying dynamic system properties 
within the MDE paradigm.