\setlength{\parindent}{1cm}
\setcounter{page}{1}
\pagenumbering{roman}

\begin{center}
  \textbf{\large{ACKNOWLEDGEMENTS}	}
\end{center}
\addcontentsline{toc}{chapter}{ACKNOWLEDGEMENTS}

% Version 2
I would like to express my deepest gratitude to my supervisor, Assoc. Prof. Dang Duc Hanh, 
for his invaluable guidance and unwavering support throughout the research and 
writing of this thesis. His expertise and dedication have been instrumental in shaping this work.

I am also grateful to the alumni and current members of 
the research group for their insightful discussions and 
constructive feedback, which greatly enriched my research.

Furthermore, I extend my thanks to the faculty members of the University 
of Engineering and Technology for their passionate teaching and for 
equipping me with the essential knowledge and skills that form 
the foundation of this thesis.

Lastly, I offer my gratitude to my family for their constant 
care, support, and encouragement. Their belief in me provided 
the motivation and stability I needed to pursue and complete this 
thesis.

Although I have endeavored to conduct this research to 
the highest standard, I recognize that limitations in my knowledge 
and experience may have led to unintentional shortcomings. 
I sincerely welcome comments and suggestions from professors and peers 
to enhance this work further.

To all who have supported me on this journey, I am profoundly grateful.

% Version 1
% Under the assignment of the Faculty of Information Technology, 
% University of Engineering and Technology - VNU, and with the approval and guidance 
% of my supervisor - Assoc. Prof. Dang Duc Hanh, I have completed the thesis entitled: 
% \textbf{\textit{"A support tool to specify and verify temporal properties in OCL"}}.

% To complete this thesis, I would like to express my sincere gratitude 
% to my supervisor - Assoc. Prof. Dang Duc Hanh for his dedicated guidance and support both in the 
% research lab and throughout the course of this thesis.

% I would also like to thank the alumni and fellow members of the research group who have
% for their support and valuable discussions throughout the research process.

% Then, I would like to express my gratitude to the faculty members of the University of
% Engineering and Technology for their dedicated and enthusiastic teaching, which has
% provided me with valuable foundational knowledge.

% Lastly, I would like to express my heartfelt gratitude to my family for their care, 
% support, and encouragement, which have provided me with the motivation 
% and favorable conditions to study and complete this thesis with peace of mind.

% Although I have made every effort to carry out this research to the best of my ability, 
% due to limitations in my knowledge and experience, 
% this thesis may still contain shortcomings that I may not have recognized. 
% I sincerely welcome any comments and suggestions from professors 
% and fellow students to help improve this work.

% I would like to extend my deepest thanks.
