\chapter*{APPENDIX}
\addcontentsline{toc}{chapter}{APPENDIX}

\section*{Appendix A: Summarization Examples}

This appendix reproduces one representative MoA run on the
\textit{tienphong.vn} article \texttt{nu-sinh-lop-8-nghe-an}
(\textit{Eighth-grade girl in Nghệ An assaulted on the way home from
school}, a current-affairs article). For each of the three proposer
drafts and the fused output, the full Vietnamese text is shown below.

\textbf{Source article (truncated, $\sim 480$ tokens).}\;
\textit{Ngày 12 tháng 5, Công an huyện Diễn Châu (tỉnh Nghệ An) đã
khởi tố vụ án hình sự đối với một nhóm nam thanh niên có hành vi
cố ý gây thương tích nghiêm trọng đối với một nữ sinh lớp 8 trên
đường đi học về \ldots\ (full text in
\texttt{metrics\_reports/.../full-article.txt}).}

\textbf{Proposer draft 1 (GPT-4o-mini).}\;
\textit{Một nữ sinh lớp 8 ở Nghệ An bị hành hung trên đường đi học về,
khiến cộng đồng dân cư bức xúc. Công an huyện Diễn Châu đã khởi tố
vụ án và xác định nhóm thanh niên gây án. Gia đình nạn nhân yêu cầu
xử lý nghiêm minh.}

\textbf{Proposer draft 2 (Claude Haiku 4.5).}\;
\textit{Công an huyện Diễn Châu (Nghệ An) khởi tố vụ án cố ý gây
thương tích sau khi một nữ sinh lớp 8 bị nhóm nam thanh niên tấn
công trên đường đi học về vào chiều ngày 11/5. Nạn nhân hiện đang
được điều trị tại bệnh viện huyện; cơ quan điều tra đã triệu tập
các đối tượng liên quan.}

\textbf{Proposer draft 3 (Gemini 2.5 Flash).}\;
\textit{Vào chiều 11/5/2026, một nữ sinh lớp 8 trên địa bàn xã ABC
(huyện Diễn Châu, Nghệ An) đã bị một nhóm nam thanh niên hành hung
khi đang trên đường đi học về. Sau khi tiếp nhận trình báo, Công an
huyện đã khởi tố vụ án; lực lượng chức năng đang truy tìm các đối
tượng còn lại.}

\textbf{Fused output (GPT-4o aggregator).}\;
\textit{Chiều 11 tháng 5, một nữ sinh lớp 8 trên địa bàn huyện Diễn
Châu, tỉnh Nghệ An, đã bị một nhóm nam thanh niên hành hung trên
đường đi học về. Ngày hôm sau, Công an huyện Diễn Châu khởi tố vụ án
cố ý gây thương tích và triệu tập các đối tượng liên quan; nạn nhân
hiện được điều trị tại bệnh viện huyện. Gia đình nạn nhân và đại
diện chính quyền địa phương đề nghị cơ quan tố tụng xử lý nghiêm
hành vi tấn công học sinh, đồng thời tăng cường an ninh đoạn đường
học sinh đi lại.}

\textbf{Observation.} The fused output (i)~retains the temporal anchor
(11/5) that Claude and Gemini supplied but GPT-4o-mini omitted;
(ii)~retains the legal action (\textit{khởi tố}) that all three
proposers supplied; (iii)~adds the institutional context (the call by
the family and local authorities to strengthen security on school
routes) that is in the source article but absent from all three
drafts. The Coverage gain ($+0.16$ in Section~4.3.1) is the
arithmetic average of effects like these across $50$ articles.

\section*{Appendix B: Fact-Checking Examples}

This appendix reproduces the fact-check pipeline on the same article.

\begin{itemize}
  \item \textbf{Claim 1} --- ``Một nữ sinh lớp 8 bị một nhóm nam
        thanh niên hành hung trên đường đi học về.''
        \\Tavily top result: \textit{tuoitre.vn} report on the same
        incident, dated $11/5/2026$. Verdict: \texttt{entailed},
        trust score $90$.
  \item \textbf{Claim 2} --- ``Công an huyện Diễn Châu đã khởi tố
        vụ án cố ý gây thương tích.''
        \\Tavily: \textit{vnexpress.net} legal-news section, same
        event. Verdict: \texttt{entailed}, trust score $85$.
  \item \textbf{Claim 3} --- ``Các đối tượng đã bị triệu tập.''
        \\Tavily: police-press-release excerpt mirrored by
        \textit{thanhnien.vn}. Verdict: \texttt{entailed}, trust
        score $80$.
  \item \textbf{Claim 4} --- ``Đại diện chính quyền địa phương đề
        nghị tăng cường an ninh đoạn đường học sinh đi lại.''
        \\Tavily: top results do not directly quote a local-government
        request along this exact line. Verdict:
        \texttt{not\_mentioned}, trust score $40$.
\end{itemize}

\noindent Aggregated trust score: $0.5 \cdot \text{entailed}\,\% \;+\;
0.5 \cdot \text{avg trust} = 74$. UI label: \textit{Unverified ---
check sources}. The \texttt{verified} boolean is \texttt{false}.
This example illustrates the design intent of the trust-score
threshold: a single \texttt{not\_mentioned} claim out of four is
enough to drop the article out of the \texttt{verified} band, even
though no claim is \textit{contradicted}.

\section*{Appendix C: Full Evaluation Results Table}

The full per-article evaluation table is exported as
\texttt{fusion\_reports/results/unified-report-2026-05-09T11-21-44-072Z.md}
and the accompanying JSON sidecar. The MD file lists, for each
approach, the $n$ rows used, the per-metric mean, and the sign-test
$p$-value where applicable. The complete file is too long to include
inline; the headline tables are reproduced in Chapter~4
(Tables~4.1--4.5).

\section*{Appendix D: Unified Three-Axis Report}

The unified three-axis report (the canonical artefact summarised in
Chapter~4) is produced by
\texttt{npm run report:unified} and is committed at the path above.
Optional flags:

\begin{itemize}
  \item \texttt{--since 2026-04-01} --- restrict to rows after a date.
  \item \texttt{--task-ids <uuid1>,<uuid2>} --- restrict Axis~C to a
        specific human-eval task list.
  \item \texttt{--json} --- write a machine-readable sidecar alongside
        the Markdown report.
\end{itemize}

\section*{Appendix E: Aggregator Prompt Engineering Ablation}

The three aggregator-prompt variants discussed in Section~3.4.5 and
Section~4.2.2 are reproduced verbatim below. Each variant was run on
the same $50$ articles; the summary metrics are reported in the
CLAUDE.md three-way comparison table.

\textbf{Variant: \texttt{baseline} (no source in aggregator
context).}\;
``Bạn là một biên tập viên báo chí Việt Nam. Dưới đây là $n$ bản tóm
tắt cho cùng một bài báo \ldots\ Hãy tổng hợp thành một bản tóm tắt
cuối cùng, trung lập, văn phong báo chí.''

\textbf{Variant: \texttt{v1-strict-source} (source injected with
strict rules).}\;
Adds: ``KHÔNG được thêm thông tin không có trong bài báo gốc dưới đây
\ldots\ \texttt{<article>...</article>}.''

\textbf{Variant: \texttt{v2-soft-source} (source injected with soft
wording, shipped).}\;
Adds: ``Bài báo gốc được cung cấp dưới đây để tham khảo
\ldots\ \texttt{<article>...</article>}.'' Plus the
post-\texttt{afac46e} removal of the phrase ``cô đọng''.

\textbf{Finding (re-stated).} The mere \textit{presence} of the
source article in the aggregator's context shifts behaviour from
draft-stitching to editorial synthesis. The exact rule wording
(\texttt{v1} strict vs.\ \texttt{v2} soft) does not meaningfully move
the metrics. The shipped variant is \texttt{v2} on aesthetic grounds:
softer wording produced fewer adversarial edge cases when the source
itself contained quotations or hedged claims.

\section*{Appendix F: Installation Guide}

\textbf{Prerequisites.} Node.js $\geq 18$, Python $\geq 3.10$,
\texttt{git}, and accounts for OpenAI, Anthropic, Google (Gemini),
Tavily, Supabase, and Hugging Face Spaces.

\textbf{Backend.}

\begin{enumerate}
  \item \texttt{cd backend \&\& npm install}
  \item Create \texttt{backend/.env} with the variables listed in
        the project README:
        \texttt{OPENAI\_API\_KEY}, \texttt{TAVILY\_API\_KEY},
        \texttt{NEXT\_PUBLIC\_SUPABASE\_URL},
        \texttt{NEXT\_PUBLIC\_SUPABASE\_ANON\_KEY},
        \texttt{SUPABASE\_SERVICE\_ROLE\_KEY},
        \texttt{BERT\_SERVICE\_URL} (optional).
  \item Apply Supabase migrations from
        \texttt{backend/supabase/migrations/}.
  \item \texttt{npm run dev} --- listens on
        \texttt{http://localhost:3000}.
\end{enumerate}

\textbf{Extension.}

\begin{enumerate}
  \item \texttt{cd extension \&\& npm install}
  \item Optional: create \texttt{extension/.env} with
        \texttt{PLASMO\_PUBLIC\_API\_URL=http://localhost:3000/api}.
  \item \texttt{npm run dev} --- Plasmo opens a Chrome dev profile
        with the extension loaded; hot-reload is on.
  \item \texttt{npm run package} produces a zip for distribution.
\end{enumerate}

\textbf{BERTScore microservice.}

\begin{enumerate}
  \item \texttt{cd bert \&\& pip install -r requirements.txt}
  \item \texttt{uvicorn main:app --host 0.0.0.0 --port 7860}
  \item For production: push the \texttt{bert/} directory to a Hugging
        Face Space with a Docker SDK; set
        \texttt{BERT\_SERVICE\_URL} in the backend \texttt{.env} to
        the Space URL.
\end{enumerate}

\textbf{Reproducing the experiments.}

\begin{itemize}
  \item Batch with judge enabled:
        \\\texttt{npx tsx output-fusion/scripts/collect-metrics.ts \textbackslash}
        \\\texttt{\quad --input output-fusion/scripts/sample-urls-tienphong-50.json \textbackslash}
        \\\texttt{\quad --judge-mode both --judge-style rubric --judge-model gpt-4o}
  \item Unified report:
        \\\texttt{npm run report:unified -- --json}
  \item P0-8 (fused vs.\ gpt-4o-alone):
        \\\texttt{npx tsx output-fusion/scripts/run-single-baseline.ts \textbackslash}
        \\\texttt{\quad --input output-fusion/scripts/sample-urls-tienphong-50.json}
        \\\texttt{npx tsx output-fusion/scripts/compare-fused-vs-single.ts}
  \item Axis~C: build tasks at \texttt{/evaluate/admin}, share the
        printed \texttt{/evaluate?task=<uuid>} links with raters,
        view aggregates at \texttt{/evaluate/admin} (Review tab).
\end{itemize}
