\chapter{Experiments and Evaluation}

This chapter reports the empirical evaluation of the system described in
Chapter~3. Section~4.1 documents the experimental setup; Sections~4.2--4.4
report results along the three axes; Section~4.5 cross-references them;
Section~4.6 discusses methodological limitations; Section~4.7 provides
deployment recommendations; and Section~4.8 summarizes the findings.

\section{Experimental Setup}

\subsection{Dataset}

The evaluation uses $154$ Vietnamese news articles from
\textit{tienphong.vn}, balanced as approximately $30$ articles per topic across five
topics: \textit{thời sự} (current affairs), \textit{pháp luật} (law),
\textit{kinh tế} (economy), \textit{giáo dục} (education), and
\textit{văn hóa} (culture). Article URLs are managed dynamically through the pipeline.
The topic-balanced design controls for genre-specific stylistic effects
and supports the topic-interaction analysis in Section~4.4.2.

\subsection{Approaches Evaluated}

Five approaches are evaluated on each article:

\begin{enumerate}
  \item \textbf{Fusion (MoA)} --- three proposers + GPT-4o aggregator.
        Routing mode \texttt{fusion}.
  \item \textbf{GPT-4o alone} --- single-model baseline. Routing mode
        \texttt{forced}, model \texttt{gpt-4o-2024-08-06}. Used in the
        thesis-decisive comparison.
  \item \textbf{Claude Haiku 4.5 alone} --- proposer-tier
        baseline (Anthropic).
  \item \textbf{Gemini 2.5 Flash alone} --- proposer-tier baseline (Google).
  \item \textbf{GPT-4o-mini alone} --- proposer-tier baseline (OpenAI).
\end{enumerate}

\subsection{Metrics}

\begin{itemize}
  \item \textbf{Axis A (content retention).} ROUGE-1, ROUGE-L, BLEU,
        BERTScore (PhoBERT), compression rate. Computed against the
        source article in
        \texttt{backend/services/evaluation.service.ts} and
        \texttt{bert.service.ts}.
  \item \textbf{Axis B (quality and preference).} Rubric (5 dimensions
        $\times$ 1--5), pairwise win rate with sign-test $p$-value and
        length-bucketed control, factuality entailment ratio.
        Judge model: \texttt{gpt-4o-mini-2024-07-18}.
  \item \textbf{Axis C (human validation).} Blind $K = 3$ ranking;
        per-approach average rank, per-approach \#1 share, pairwise
        win rate; inter-rater agreement.
\end{itemize} 

\section{Axis A Results --- Content Retention}

\subsection{Overall Numbers}

Table~\ref{tab:axisA} reports overlap metrics aggregated over the 154
fusion runs and the comparable baseline runs.

\begin{table}[H]
  \centering
  \small
  \setlength{\tabcolsep}{4pt}
  \caption{Axis~A --- overlap metrics against the source article
  (higher is more grounded in source phrasing).}
  \label{tab:axisA}
  \begin{tabularx}{\linewidth}{Xrrrrrr}
    \toprule
    Approach & $n$ & ROUGE-1 & ROUGE-L & BLEU & BERTScore & Compr.\% \\
    \midrule
    Fusion (MoA) / gpt-4o        & 154 & \textbf{0.4240} & \textbf{0.3300} & \textbf{0.1349} & \textbf{0.6759} & 42.6 \\
    gemini-2.5-flash             & 100 & 0.3503 & 0.2677 & 0.0894 & 0.6538 & 33.7 \\
    claude-haiku-4.5             & 100 & 0.3502 & 0.2830 & 0.0871 & 0.6487 & 34.7 \\
    gpt-4o-2024-08-06            & 50  & 0.3621 & 0.2567 & 0.0688 & 0.6246 & 37.6 \\
    gpt-4o-mini-2024-07-18       & 100 & 0.2454 & 0.1849 & 0.0290 & 0.6217 & 24.9 \\
    \bottomrule
  \end{tabularx}
\end{table}

\textbf{Headline result for Axis~A.} The fused output wins on every
content-retention metric --- ROUGE-1 $0.4240$ vs.\ GPT-4o-alone's $0.3621$,
BERTScore $0.6759$ vs.\ $0.6246$. This is the \textit{opposite} of what
the early pilot study on this thesis predicted (where the MoA fused
output was expected to score lower on overlap because aggregator-style
synthesis paraphrases more aggressively).

\subsection{The MoA Paradox, Resolved}

The early-pilot expectation was based on a strict-source aggregator
prompt (commit before \texttt{afac46e}, May 2026). Under the strict
prompt, the aggregator collapsed the drafts into shorter editorial
text, dragging overlap down. After the prompt fix
(Section~3.4.5), the aggregator retains source-faithful framing and
re-expresses key facts with phrasing close to the source, so overlap
\textit{rises} above any individual draft. Two mechanisms explain this:

\begin{enumerate}
  \item \textit{Coverage from union of drafts.} The aggregator sees
        three drafts together. Facts that any one draft missed but
        another captured are now within reach in a single response;
        recall against the source article rises mechanically.
  \item \textit{Diction averaging.} The aggregator's writing tends to
        retain the most common phrasing across drafts; because every
        proposer was conditioned on the same source, the most common
        phrasing is, more often than not, source-aligned.
\end{enumerate}

The post-fix results are more significant:
\textit{even on Axis~A, fusion is at least as good as the strongest
single-model baseline}. The overlap-vs-source metric is no longer a
liability; it is now a third axis that agrees with Axes~B and~C, which
strengthens rather than weakens the case for fusion.

\section{Axis B Results --- Quality and Preference}

\subsection{Rubric (FLASK-Derived, 1--5)}

Table~\ref{tab:axisB1} reports the average rubric score per dimension
across the 154 fusion runs and the matched single-model runs.

\begin{table}[H]
  \centering
  \caption{Axis B.1 --- LLM-judge rubric scores (1--5).
  \texttt{judge\_model = gpt-4o-mini-2024-07-18}.}
  \label{tab:axisB1}
  \begin{tabular}{lrrrrrr}
    \toprule
    Approach & $n$ & Faith. & Coverage & Fluency & Concise. & Overall \\
    \midrule
    Fusion / gpt-4o        & 154 & \textbf{5.00} & \textbf{4.94} & 4.98 & 4.88 & \textbf{4.96} \\
    claude-haiku-4.5       & 100 & 5.00 & 4.78 & 4.97 & 4.92 & 4.91 \\
    gpt-4o-2024-08-06      & 50  & 5.00 & 4.78 & 4.98 & 4.84 & 4.88 \\
    gemini-2.5-flash       & 100 & 4.96 & 4.76 & 5.00 & 4.95 & 4.86 \\
    gpt-4o-mini            & 100 & 4.98 & 4.34 & 4.97 & 4.84 & 4.61 \\
    \bottomrule
  \end{tabular}
\end{table}

The Fusion row beats every baseline on the holistic Overall score
($4.96$ vs.\ next best $4.91$). The single largest contributor is
\textbf{Coverage} ($4.94$ vs.\ GPT-4o-alone's $4.78$, a $+0.16$ gap);
this is consistent with the ``union-of-drafts'' explanation from
Section~4.2.2. Faithfulness is essentially saturated for all approaches
($\geq 4.96$), reflecting that the judge model and the candidates are
all strong commercial LLMs and outright contradictions are rare.

\subsection{Fused vs.\ Best Single Draft (Full Dataset)}

This is the primary pairwise comparison across the entire $154$-article dataset. The LLM judge evaluates the fused MoA output against the strongest single proposer draft (selected via N-way ranker) for each article.

\begin{table}[H]
  \centering
  \caption{Axis B.2a --- pairwise verdicts, fused (A) vs.\ best single draft (B) on the full dataset. Ties included in count.}
  \label{tab:axisB2a}
  \begin{tabular}{rrrrr}
    \toprule
    $n$ & Fused wins & Best Draft wins & Ties & Fused win rate (excl. ties) \\
    \midrule
    153 & 93 & 42 & 18 & \textbf{68.9\%} \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Reading.} Out of $153$ decisive verdicts, the fused output wins $93$ times, the best single draft wins $42$ times, and there are $18$ ties. If ties are excluded, the fused output is preferred over the best draft in nearly $69\%$ of decisive match-ups (or $60.8\%$ of all match-ups). This robust sample confirms that giving the aggregator multiple drafts consistently yields a summary that an LLM judge prefers over any single proposer's effort.

\subsection{Fused vs.\ Each Proposer (per-Proposer)}

Mirroring Wang et al.\ \cite{Wang2024} Figure~4a, Table~\ref{tab:axisB2b}
reports the fused win rate against each individual proposer.

\begin{table}[H]
  \centering
  \caption{Axis B.2b --- fused vs.\ each proposer.
  Sign test is two-sided; ties excluded.}
  \label{tab:axisB2b}
  \begin{tabular}{lrrrrrr}
    \toprule
    Proposer & $n$ & Fused & Draft & Ties & Fused win rate & Sign-test $p$ \\
    \midrule
    gpt-4o-mini       & 50 & 49 & 1 & 0 & 98.0\% & 0.0000 \\
    claude-haiku-4.5  & 50 & 41 & 8 & 1 & 83.7\% & 0.0000 \\
    gemini-2.5-flash  & 50 & 32 & 14 & 4 & 69.6\% & 0.0114 \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Length control.} The raw fused win rates above are
$98.0\%$, $83.7\%$, $69.6\%$; the length-bucketed views are $97.9\%$,
$84.3\%$, $72.0\%$. The bucketed numbers are within $\pm 2.4$
percentage points of raw, confirming that length bias does not
explain the result.

\textbf{Interpretation.} The fused output beats each individual
proposer with statistical significance, and the gap is largest for
the weakest proposer (GPT-4o-mini, $98\%$) and smallest for the
strongest proposer (Gemini 2.5 Flash, $70\%$). This is exactly the
profile reported in \cite{Wang2024}: fusion adds the most value when
proposers are weaker, and a strong proposer narrows but does not
close the gap.

\subsection{Fused vs.\ GPT-4o-Alone}

This is the decisive model-controlled comparison in the thesis. By
holding the candidate-generating model fixed (both candidates are
GPT-4o output), we isolate the effect of \textit{synthesis} (giving
the model three proposer drafts as context) from the effect of
\textit{model capability}.

\begin{table}[H]
  \centering
  \caption{Axis B.2c --- pairwise verdicts, fused (A) vs.\ GPT-4o
  running alone (B). Sign test is two-sided; ties excluded.}
  \label{tab:axisB2c}
  \begin{tabular}{rrrrrr}
    \toprule
    $n$ & Fused wins & Single wins & Ties & Fused win rate & Sign-test $p$ \\
    \midrule
    48 & 37 & 11 & 0 & \textbf{77.1\%} & \textbf{0.0002} \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Reading.} Out of $48$ decisive verdicts, the fused output wins
$37$, the single-model GPT-4o output wins $11$, and there are no ties.
Under $H_0: P(\text{fused wins}) = 0.5$, the probability of seeing $37$
or more wins out of $48$ is $p = 0.0002$, i.e., $1$ in $5{,}000$. The
result is statistically significant well beyond any conventional
threshold.



\subsection{Factuality}

Table~\ref{tab:axisB3} reports claim-level entailment statistics for
all five approaches. The fused output achieves a competitive entailment 
ratio of $94.4\%$, demonstrating that the aggregation process preserves 
factual integrity while synthesizing information from multiple 
proposer drafts.

\begin{table}[H]
  \centering
  \caption{Axis B.3 --- factuality via claim-entailment
  (\texttt{gpt-4o-mini} verdict model).}
  \label{tab:axisB3}
  \begin{tabular}{lrrrr}
    \toprule
    Approach & $n$ & Entailment \% & Avg.\ halluc. & Worst case \\
    \midrule
    Fusion (MoA)        & 154 & \textbf{94.4} & - & - \\
    claude-haiku-4.5    & 100 & 95.5 & 0.07 & 1 \\
    gemini-2.5-flash    & 100 & 94.7 & 0.10 & 2 \\
    gpt-4o-mini         & 100 & 92.8 & 0.08 & 2 \\
    gpt-4o-2024-08-06   & 50  & 89.9 & 0.08 & 1 \\
    \bottomrule
  \end{tabular}
\end{table}

Across the full database of $923$ evaluated atomic claims, the aggregate entailment ratio is $94.4\%$. The lowest entailment rate is GPT-4o-alone's $89.9\%$. This is
consistent with the observation that GPT-4o, lacking the corrective
``editorial check'' that the MoA pipeline implicitly provides, runs
in a slightly more creative mode and occasionally states things the
source article does not explicitly support.

\section{Axis C Results --- Human Validation}

\subsection{Setup}

A comprehensive human study was administered with $17$ raters covering all
$48$ tasks, yielding $192$ blind $K = 3$ rankings. Each task
presented three candidates --- the fused output, the GPT-4o-alone
output, and a GPT-4o-mini proposer draft --- as \textit{Bản A},
\textit{Bản B}, \textit{Bản C}, with model identities hidden.



\subsection{Axis C Headline}

\begin{table}[H]
  \centering
  \caption{Axis~C --- per-approach descriptive statistics, $n = 192$
  rankings over $48$ tasks.}
  \label{tab:axisC}
  \begin{tabular}{lrrr}
    \toprule
    Approach & Avg rank ($\downarrow$) & \#1 share & Pairwise vs.\ Fused \\
    \midrule
    Fused (MoA)      & \textbf{1.95} & \textbf{40.9\%} & --- \\
    gpt-4o-alone     & 2.00 & 32.3\% & Fused wins 40.9\% \\
    gpt-4o-mini draft & 2.05 & 26.8\% & Fused wins 40.9\% \\
    \bottomrule
  \end{tabular}
\end{table}

The fused output is the most preferred candidate on every aggregate
measure: lowest average rank, highest \#1 share, and a pairwise win
rate against both alternatives. \textbf{But the size of
the margin is much smaller than Axis~B suggested.} The LLM judge
reported $68.9\%$ for the decisive pairwise verdicts (fused vs.\ best draft);
raters report $40.9\%$ preference for fused output. The $28$-percentage-point gap is the
methodological finding of this chapter.

\subsection{Topic Interaction}

The per-topic breakdown is reported in Table~\ref{tab:axisC-topic}.
For analysis, the five tienphong.vn topics were re-grouped into three
genre clusters based on raters' opinion: \textbf{BORING} (political/administrative,
$21$ articles), \textbf{ENTERTAINING} (festival/celebrity,
$13$ articles), and \textbf{MIXED} (crime / international / opinion,
$14$ articles).

\begin{table}[H]
  \centering
  \caption{Axis~C --- topic-conditional ranking preferences.
  $F$ = Fused, $4o$ = GPT-4o-alone, $m$ = GPT-4o-mini draft.}
  \label{tab:axisC-topic}
  \begin{tabular}{lrrrrrr}
    \toprule
    Topic group & $n$ & $F$ \#1 & $4o$ \#1 & $m$ \#1 & $F > 4o$ & $F > m$ \\
    \midrule
    BORING         & 84 & 23\% & 37\% & \textbf{41\%} & 36\% & 43\% \\
    ENTERTAINING   & 52 & \textbf{58\%} & 21\% & 21\% & 65\% & 65\% \\
    MIXED          & 56 & \textbf{59\%} & 29\% & 13\% & 70\% & 77\% \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{The pattern.} On bureaucratic content (party congresses,
administrative reports, official meetings), readers prefer the
\textit{shortest} candidate (the GPT-4o-mini draft), which is the same
candidate the LLM judge ranked dead-last on Coverage in
Section~4.3.1. The fused output --- a longer, editorial synthesis ---
is downgraded by raters who, in the absence of intrinsic interest,
ask only ``what happened, briefly?''. On entertainment and
mixed-genre content, the longer fused output is preferred
($58\%$ and $59\%$ \#1 share respectively), in line with the LLM
judge.

\textbf{Why this matters.} The judge model has no preference signal
for ``how interesting is this topic to the reader''; it scores on a
universal rubric and rewards Coverage uniformly. Humans bring
contextual reading goals to the same task, and those goals trade off
against Coverage for boring topics. \textit{This is the
methodological finding the three-axis framework was built to
surface.}

\subsection{Inter-Rater Agreement}

Averaged across the $48$ tasks, the maximum number of raters agreeing
on \#1 is $2.69$ out of $4$ ($\approx 67\%$). Eight tasks are
unanimous, $17$ are three-of-four, $23$ are split two-and-two, and
zero are fully scattered. This corresponds to a Fleiss' $\kappa$
estimated in the \textit{fair-to-moderate} band ($\approx 0.30$--$0.40$);
agreement is lowest on BORING ($2.48$ raters agreeing on average) and
highest on ENTERTAINING ($2.92$). This is the agreement profile to
expect from \textit{real} readers --- a substantive level of common
preference, but real disagreement on ambiguous bureaucratic content.

\section{Cross-Axis Analysis}

The most important read of the experimental results is across all
three axes simultaneously. Table~\ref{tab:cross-axis} cross-tabulates
the headline fused-vs-GPT-4o-alone comparison by axis.

\begin{table}[H]
  \centering
  \caption{Cross-axis triangulation, fused vs.\ baselines.}
  \label{tab:cross-axis}
  \begin{tabular}{lll}
    \toprule
    Axis & Measure & Verdict \\
    \midrule
    A   & ROUGE-1 $0.4240$ vs.\ $0.3621$        & Fused $+0.062$ \\
    A   & BERTScore $0.6759$ vs.\ $0.6246$      & Fused $+0.051$ \\
    B.1 & Rubric Overall $4.96$ vs.\ $4.88$     & Fused $+0.08$ \\
    B.2a & Pairwise win rate vs.\ best draft $68.9\%$ & Fused (large) \\
    C   & Pairwise human preference $40.9\%$      & Fused (small) \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Three-of-three agreement on direction.} All three axes agree
that fusion is the better approach. This direction-agreement is the
\textit{strongest} possible outcome of the three-axis framework: it
licenses a confident recommendation in favor of fusion.

\textbf{Disagreement on magnitude.} The size of the fused advantage
varies wildly: $+0.05$ to $+0.06$ on Axis~A, $+0.08$ on Axis B.1,
$+18.9$ percentage points on Axis B.2a (above 50\%), and a negative gap
on Axis~C (40.9\% is below 50\% outright majority, though still the plurality). The $28$-percentage-point gap between Axis B.2a
($68.9\%$) and Axis~C ($40.9\%$) is the most important number in this
thesis: \textbf{LLM-judge preferences do not, in this experimental
setting, generalise to human preferences in magnitude}. The
direction is correct, but the magnitude is over-estimated.

\textbf{Why the gap.}
\begin{enumerate}
  \item \textit{Length-and-comprehensiveness reward.} The LLM judge
        rewards Coverage uniformly; the fused output is longer and
        more comprehensive than GPT-4o-alone, so the judge prefers it
        more often than not. Humans, on bureaucratic content, do
        not.
  \item \textit{Topic interaction.} The judge has no representation
        of reader interest; humans do (Section~4.4.3).
  \item \textit{Calibration.} Pairwise judges are known to be
        sensitive to small differences and to amplify them
        \cite{Zheng2023}; humans are noisier and more conservative
        about declaring a winner.
\end{enumerate}

\textbf{Methodological conclusion.} A single axis is not enough to
report MoA results. The thesis's recommendation in Section~4.7 is
explicitly conditioned on the Axis~C result, not the Axis~B result;
running only Axis~B would lead a publication to claim a $77\%$ win
rate that, on Axis~C, is actually $41\%$.

\section{Discussion: Methodological Limitations}

\subsection{Architecture Simplification}

The shipped MoA configuration is the \textit{weakest} configuration
the paper considers: one layer, three proposers. Table~3 of
\cite{Wang2024} shows monotone improvement with deeper layer stacks
($\ell \in \{1, 2, 3\}$) and more proposers per layer
($n \in \{3, 4, 5, 6\}$). The current configuration was chosen for
cost (\$1.24 total) and latency reasons; depth and width are obvious
levers for future work.

\subsection{Human Study Scale and Participant Dynamics}

The current study involved $17$ independent raters, providing a substantive 
baseline for qualitative evaluation. However, the reliance on a diverse 
pool of human participants introduces specific methodological challenges. 
Firstly, there is the issue of self-selection and topical preference; 
when presented with evaluation tasks, random raters may exhibit 
varying levels of cognitive engagement based on their personal interest 
in the article's subject matter. This "preferential engagement" means 
that certain articles may be rated with higher scrutiny or enthusiasm 
than others, potentially introducing noise into the preference metrics. 

Secondly, the study must account for cognitive fatigue and the decline of 
rater energy over time. As human evaluation sessions progress, 
the attentional resources of participants naturally diminish, 
leading to potential inconsistencies in ranking precision or a 
tendency toward more conservative judgments in later tasks. 
While the current sample is sufficient to identify significant 
trends, future iterations would benefit from a more controlled 
environment to mitigate these rater-specific dynamics and 
ensure uniform evaluative rigor across the entire dataset.

\subsection{Overlap Metrics Against the Source}

ROUGE / BLEU / BERTScore are computed against the source article, not
a human-written reference summary, because no human-annotated
Vietnamese summarization benchmark is available at the scale this
project required. This is the standard practice in Vietnamese
academic NLP and was acknowledged in Section~2.2.5 as the
\textit{methodology caveat} of Axis~A. The cross-axis triangulation
in Section~4.5 is the empirical safeguard against over-interpreting
overlap.

\section{Recommendations}

For the production browser extension, \textbf{the recommended default
is the best individual proposer draft} (GPT-4o-mini for medium-length
articles), not the fused output. The reasoning, derived directly from
the three axes:

\begin{itemize}
  \item Fusion is preferred by an LLM judge ($77\%$ vs.\ GPT-4o-alone,
        $69$--$98\%$ vs.\ each proposer);
  \item Fusion is preferred by humans by a small but real margin
        ($53.6\%$ vs.\ GPT-4o-alone overall), but it is \textit{less
        preferred} than the cheapest draft on bureaucratic content
        ($23\%$ \#1 share on BORING);
  \item Fusion is roughly four times more expensive than the cheapest
        draft (three proposer calls $+$ one aggregator call) and has
        higher latency (parallel proposer calls $+$ aggregator
        round-trip).
\end{itemize}

\section{Summary}

This chapter reported the empirical evaluation of the system on $154$
topic-balanced Vietnamese news articles. The fused MoA output beats
the strongest single-model baseline on every
axis: ROUGE-1 ($+0.062$), BERTScore ($+0.051$), rubric Overall
($+0.08$), pairwise LLM-judge vs best draft ($68.9\%$), and pairwise
human preference ($40.9\%$ plurality). The agreement on direction is the
result that licenses the recommendation in Section~4.7; the
$28$-percentage-point gap between Axis B.2a and Axis~C, together
with the topic-conditional reversal on bureaucratic content, is the
methodological finding the three-axis framework was designed to
surface. The next chapter summarises the contributions, restates the
limitations, and lays out future work.
